---
title: "Final Project - The witch trials of Scotland"

output:
  word_document: default
  html_document: default
---
<br> 

# Introduction 

The Scottish witchcraze began in earnest in 1590, with the trial of a group of people, mainly women, from East Lothian. They were accused of meeting with the devil and conjuring up storms to destroy James VI on his return from Denmark with his bride, Anne. The King, who personally examined the accused, composed his own treatise on the subject, Daemonologie. The trial of the North Berwick coven began a pattern of persecution of suspected witches by torture and capital punishment. Witch hunts often took place at times of upheavel, warfare, famine or diasease. In Scotland the main phases of the witchcraze were the late 1620s, the late 1640s and in the begining of the 1660s. In the late 16th- and 17th-century Scotland, between three and four thousand people were tortured and executed as witches, a group identified as threatening social stability. The methods of torture involved devices such as thumbscrews with padlock, iron gags, sleep deprivation, burning feets and much more. Most of those accused were women - spinters or widows without means of support and unable to defend themselves. Often they had a local reputation for herbal remedies, folk medicine and healing. The persecution of witches in Scotland was an extreme example of a movement which swept Europe. 

<br>

This project seeks to look into the torture methods used in the during the witchcraze in Scotland. By using R we will be able to tell the most frequent torture methods and the development in torture methods used through the time period our chosen dataset provides. 

To do so I have created a proof of concept and guidelines explaining every step on the way.
<br>

# Collecting packages

First of you must collect the package "tidyverse." The "tidyverse" package collects some of the most versatile R packages: ggplot2, dplyr, tidyr, readr, purrr, and tibble. The packages work in harmony to clean, process, model, and visualize data. Tidyverse is installed by the following steps:

- Click on Packages 
- Click install 
- Choose Tidyverse 
- Install 

You have now collected the package "tidyverse." 

Then you must collect the package "Hmisc" short for Harrell Miscellaneous which contains many functions useful for data analysis, high-level graphics, utility operations, functions for computing sample size and power, importing and annotating datasets, imputing missing values, advanced table making, variable clustering, character string manipulation, conversion of R objects to LaTeX and html code, and recoding variables. This packages is needed as we want to import an access database into R to filter out what we need to make a statistical analysis. As the dataset is an mdb file we need to load the package "Hmisc" as it has the mdb.get()-fuction that we need to extract the data from the mdb-file. As I am a mac user loading the "Hmisc" package was not as easy as the tidyverse package. Help was provided by stackoverflow.com and osxdaily.com

<br>

To load the "Hmisc" package use following steps:

- Install the Hmisc packages for R (like you did with tidyverse)
- Install mdbtools version 0.7.1 by homebrew by typing the following into the "Terminal" application: brew install mdbtools 
- To install homebrew on Mac OS go to the "Terminal" application and run the following command into a single line of the terminal: 

/usr/bin/ruby -e "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install)" 

- Follow the instructions in "Terminal" 
- Enter the administrator password to begin installation. When complete the message "Installation complete!" will appear in Terminal 

<br> 
Both tidyverse and Hmisc has now been collected and we are ready to explore the dataset. 

```{r message=FALSE}
library(tidyverse)
library(Hmisc)
```
# Collecting database

Now we use the mdb.get()-function in the Hmisc package to read the dataset into R. 
```{r}
d <- mdb.get("./data/Witchcraftsurvey_download.mdb")
```

<br>
To show the contents of the datebase we use the command contents(d) 
```{r}
contents(d)
```

We are now able to see that the dataset contains 38 tables with data. 
<br>

As we are interested in analysing the torture methods we need to extract the table with torture methods into its own dataframe by using the following command:
```{r}
Torture <- mdb.get("data/Witchcraftsurvey_download.mdb" , tables = "WDB_Torture" )
```
<br>

# Visualisation of torture methods 
<br> 
To make a visualisation of the torture methods I got help from hax.odsbjerg.dk which is a great tool and cheatsheet to using R. A visualasation is possible using the following command: 
```{r}
 Torture %>%
  count(Torturetype, sort = TRUE) %>%
  top_n(3000) %>%
  mutate(Torturetype= reorder(Torturetype, n)) %>%
  ggplot(aes(x = Torturetype, y = n)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
      labs(x = "Torturemethod",
      y = "Count",
      title = "Most frequent use of torture methods 1580s-1690s")
```

<br> 

# Changing torture methods 
<br> 
Before looking into the changing torture methods during the time span the dataset provides I have decided to clean up the dates in the torture data. To do so I have used Open Refine.To clean the torture dataset I have used following commands in Open Refine:

- Click the arrow on the column you want to edit 
- Click "edit cells"
- Click "transform"
- Use regular expression "value.replace" - in my case: "value.replace("00","").replace(":","") 
- You have now removed 00 and : from the chosen columns 

The clean torture dataset is exported in a comma-separated value file (csv-file). To load the clean dataset into R we use following command:

```{r}
Torturedate <- csv.get("./data/tortureclean.csv")

```
<br>
We only want to explore the torture dates compared to the torture methods so we isolate the columns by using the select-cammand:

```{r}
select(Torturedate, Torturedate.as.date, Torturetype)
```



To see if the torture methods change during the time span the dataset provides we visualise the torture dates and compare them to the torture methods. 
<br>

# Collecting data with date as character 

By collecting data with date as character instead of factors is necessary to work with the data in a visualization and comparison of the torture date and torture methods and to make the data more manageable. To do so, use following command: 

```{r}
tortureclean <- read_csv("data/tortureclean.csv",  col_types = cols(Torturedate = col_character()))
```

As only the year and not the day and month is relevant in this case we use the string command to replace the day and month and leave the year standing alone: 


```{r}
tortureclean %>% 
  mutate(Torturedate = str_replace_all(Torturedate, "^\\d+/(\\d{4})", "\\1")) %>% 
  mutate(Torturedate = as.numeric(Torturedate)) -> tortureclean
```

# Facet wrap visualization

To visualize and analyse the date we use the command facet_wrap: 
<br> 

```{r}
tortureclean %>% 
  ggplot(aes(x = Torturedate))+
  geom_histogram(binwidth = 10)+
  facet_wrap(~ Torturetype)
```

